Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-05 talk/dls2014/paper/paper.tex
add a table with the memory requirements
some clarifications

\makeatletter
\usepackage{enumitem}
\usepackage{multicol}

@@

Obviously, if we are in a shared-memory setting, this model is extremely
wasteful: at a given time, most objects are identical in all, or most,
segments.  They differ only in segments that have modified but not yet
committed the object.  So we use a trick of the Memory Management Unit
(MMU) of the CPU.  At the coarse unit of one page of memory (4096
bytes), we use remappings so that, in general, pages within segments are
\emph{shared} with each other if they would contain a fully identical
copy.  Conversely, when one segment needs to write to one object
contained in a page, the complete page is first \emph{privatised:} it is
duplicated and the copy is remapped so that it will be visible in this
segment (only).  This process is illustrated in
figure~\ref{fig:Page-Remapping}.

@@

When we need to store into our objects pointers to other objects, we
need to be careful: if we were to directly store regular addresses valid
in the local segment, then the objects' binary content would have to differ
between segments.  This would prevent any sharing.  Instead, we store
pointers as \emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start
of the segment.  A $SO$ can then be interpreted from any segment by
adding it to the base address of that segment.\footnote{An alternative
model would be to simulate threads using several independent processes
in which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the state
(file descriptors, external libraries, ...) is shared.} The result is
called a \emph{Linear Address (LA)}. This is illustrated in
figure~\ref{fig:Segment-Addressing}.

@@

\begin{lstlisting}
void stm_start_transaction()
void stm_commit_transaction()
void stm_read(object_t *obj)
void stm_write(object_t *obj)
object_t *stm_allocate(ssize_t size_rounded)
STM_PUSH_ROOT(obj)
STM_POP_ROOT(obj)
\end{lstlisting}

@@

The type \lstinline!object_t! is special as it causes the
compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
its ``address-space 256'' feature. Patches are available from authors
until inclusion into the official clang.} to make all accesses through
it relative to the $\%gs$ register.  With exceptions, nearly all
accesses to objects managed by the TM system should use this type so
that the CPU will translate the reference to the right version of the
object.

@@

This area can be seen as a continuous array of bytes that is indexed
from the start of the segment by an object's reference ($SO$) divided
by 16 (to avoid false conflicts, all objects have a size of at least 16 bytes).
Instead of just setting the byte to \lstinline!true!  if the
corresponding object was read, we set it to a \lstinline!read_version!
belonging to the transaction, which will be incremented on each
commit.  Thereby, we can avoid resetting the bytes to
\lstinline!false!  on commit and only need to do this every 255
transactions. The whole code for the barrier is easily optimisable for
compilers as well as perfectly predictable for CPUs:

@@

\cfbolz{maybe it's worth explicitly pointing out here that this is one of the
    few places where the \%gs translation must not be used}
\arigo{wrong, it is used.  maybe describe a bit more why}

@@

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \lstinline!WRITE_BARRIER!  in the object's
header and call the slow path if it is set. This flag is set either if
the object is old and comes from an earlier transaction, or if there
was a minor collection which will add the flag again on all
objects. It is never set on freshly allocated objects that still
reside in the nursery.

\cfbolz{is \%s translation used here? I guess yes, currently it's possible to misunderstand that}

@@

\cfbolz{put the slow path into a figure, if it straddles a page it's too hard figuring out indentation}
The \textbf{slow path} is shown here:

\begin{lstlisting}
void write_slowpath(SO):
	// GC part:
	list_append(to_trace, SO)
	if is_overflow_obj(SO):
		SO->flags &= ~WRITE_BARRIER
		return
	// STM part
	stm_read(SO)
	lock_idx = SO >> 4
  retry:
	if write_locks[lock_idx] == our_num:
		// we already own it
	else if write_locks[lock_idx] == 0:
		if cmp_and_swap(&write_locks[lock_idx],
					    0, our_num):
			list_append(modified_old_objects, SO)
			privatize_pages(SO)
		else:
			goto retry
	else:
		w_w_contention_management()
		goto retry
	SO->flags &= ~WRITE_BARRIER
\end{lstlisting}

First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to.

@@

For the \emph{TM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
again is a simple global array of bytes that is indexed with the $SO$
of the object divided by 16. If we already own the lock, we are done.
If someone else owns the lock, we will do a write-write contention
management that will abort either us or the current owner of the
object.  If we succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

@@

>>>>

\makeatletter
\usepackage{enumitem}
\usepackage{multicol}
\usepackage{tabularx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\usepackage{float}
\floatstyle{ruled}
\newfloat{code}{tbp}{loa}
\providecommand{\codename}{Listing}
\floatname{code}{\protect\codename}

@@

Obviously, if we are in a shared-memory setting, this model is extremely
wasteful: at a given time, most objects are identical in all, or most,
segments.  They differ only in segments that have modified but not yet
committed the object.  So we use a trick of the Memory Management Unit
(MMU) of the CPU.  At the coarse unit of one page of memory (4096
bytes), we use remappings so that, in general, pages within segments are
\emph{shared} with each other if they would contain a fully identical
copy.  Conversely, when one segment needs to write to one object
contained in a page, the complete page is first \emph{privatised:} it is
duplicated and the copy is remapped so that it will be visible in this
segment (only).  This process is illustrated in
Figure~\ref{fig:Page-Remapping}.

@@

When we need to store into our objects pointers to other objects, we
need to be careful: if we were to directly store regular addresses valid
in the local segment, then the objects' binary content would have to differ
between segments.  This would prevent any sharing.  Instead, we store
pointers as \emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start
of the segment.  A $SO$ can then be interpreted from any segment by
adding it to the base address of that segment.\footnote{An alternative
model would be to simulate threads using several independent processes
in which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the state
(file descriptors, external libraries, ...) is shared.} The result is
called a \emph{Linear Address (LA)}. This is illustrated in
Figure~\ref{fig:Segment-Addressing}.

@@

\begin{lstlisting}
void stm_start_transaction()
void stm_commit_transaction()
void stm_read(object_t *SO)
void stm_write(object_t *SO)
object_t *stm_allocate(ssize_t size_rounded)
STM_PUSH_ROOT(object_t *SO)
STM_POP_ROOT(object_t *SO)
\end{lstlisting}

@@

In the following sections, whenever we use $SO$, we go through the
address translation to get to the actual contents of an object. This
is also signified by the type \lstinline!object_t!.  This type is
special as it causes the
compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
its ``address-space 256'' feature. Patches are available from authors
until inclusion into the official clang.} to make all accesses through
it relative to the $\%gs$ register.  With exceptions, nearly all
accesses to objects managed by the TM system should use this type so
that the CPU will translate the reference to the right version of the
object.

@@

This area can be seen as a continuous, segment-local array of bytes
that is indexed with an object's reference ($SO$) divided by 16. So
that each object has its own read marker, all objects have a size of
at least 16 bytes. Otherwise there would be false conflicts when
reading from two adjacent objects that share a single marker.
Instead of just setting the byte to \lstinline!true!  if the
corresponding object was read, we set it to a \lstinline!read_version!
belonging to the transaction, which will be incremented on each
commit.  Thereby, we can avoid resetting the bytes to
\lstinline!false!  on commit and only need to do this every 255
transactions. The whole code for the barrier is easily optimisable for
compilers as well as perfectly predictable for CPUs:

@@

@@

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \lstinline!WRITE_BARRIER!  in the object's
header through its $SO$ and call the slow path if it is set. This flag
is set either if the object is old and comes from an earlier
transaction, or if there was a minor collection which will add the
flag again on all objects. It is never set on freshly allocated
objects that still reside in the nursery.

@@

\begin{code}[h]
\begin{lstlisting}
void write_slowpath(SO):
	// GC part:
	list_append(to_trace, SO)
	if is_overflow_obj(SO):
		SO->flags &= ~WRITE_BARRIER
		return
	// STM part
	stm_read(SO)
	lock_idx = SO >> 4
  retry:
	if write_locks[lock_idx] == our_num:
		// we already own it
	else if write_locks[lock_idx] == 0:
		if cmp_and_swap(&write_locks[lock_idx],
					    0, our_num):
			list_append(modified_old_objects, SO)
			privatize_pages(SO)
		else:
			goto retry
	else:
		w_w_contention_management()
		goto retry
	SO->flags &= ~WRITE_BARRIER
\end{lstlisting}

\caption{Slow path of the write barrier\label{lst:w_slow}}
\end{code}

The \textbf{slow path} is shown here in Listing~\ref{lst:w_slow}.
First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to.

@@

For the \emph{TM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
is a simple, \textbf{global} array of bytes that is indexed with the
$SO$ of the object divided by 16. Note, ``global'' here really means
it is a single array with data for all segments, there is no address
translation going on to access its elements contrary to e.g.\ the
read markers array.  If we already own the lock, we are done.
If someone else owns the lock, we will do a write-write contention
management that will abort either us or the current owner of the
object.  If we succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

@@

\begin{table}[h]
  \centering
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill} }|p{2cm}|c|c|c|}
    \hline
                        & pypy-nojit  & pypy-stm-nojit & Overhead  \\
    \hline
    btree (small)       & 3.2 /  23.8 & 23.8 / 106     & 7.4 / 4.5 $\times$ \\
    \hline
    skiplist (small)    & 3.4 / 19.1  & 23.5 / 103     & 6.9 / 5.4 $\times$ \\
    \hline
    threadworms (small) & 13.6 / 28.6 & 30.1 / 150     & 2.2 / 5.2 $\times$ \\
    \hline
    mandelbrot (small)  & 3.7 / 16.6  & 21.2 / 101     & 5.7 / 6.1 $\times$ \\
    \hline
    raytrace (small)    & 9.0 / 67.0  & 22.9 / 141     & 2.5 / 2.1 $\times$ \\
    \hline
    richards (small)    & 3.7 / 16.6  & 24.4 / 116     & 6.6 / 7.0 $\times$ \\
    \hline
  \end{tabular*}
  \caption{Memory requirements reported as average GC-memory / RSS in MiB}
\end{table}

\remi{We cut off the first second of run time to not include startup.
We run on 4 threads in order to see some privatisation of pages in STM.}
